Optimal Pricing and Informal Sharing: \\ Evidence from Piped Water in Manila

### 2018-09-17 phil_paper/jmp_wjv_v2.tex
lots of good updates

A full analysis of water usage in Manila also requires describing households that choose to use water from local vendors.  For information on households that are using water from alternative sources, I turn to the 2010 Census of Population and Housing.  Households are defined as connected to the piped network if they report using a ``faucet community water system'' for their cooking, laundry, or bathing needs.\footnote{The census was designed largely for rural populations so it does not directly ask whether households are connected to a piped water network.}  Alternatively, I define households as using from a water vendor if they report any other category, which includes deep wells, peddlers, or other sources.  I then construct similar demographic measures to the connection survey using questions in the census and compare these measures across these data sources in Table~\ref{table:pawstocensus} of Appendix~\ref{appendix:pawstocensus}.  Households that own a connection in the connection survey are similar to households connected to the piped water network in the Census data.  To construct a profile of water source choices for the full population, I randomly sample from the pool of households using from vendors in 8 regions of the city so that the data form a representative sample when combined with households covered by the connection survey in each small area.\footnote{Regions are chosen to overlap with village administrative/political boundaries while ensuring that I am able to obtain a large enough sample of vendor households for each area to perform the counterfactual analysis in Section~\ref{section:fixedcostestimation}.}

@@

With the full set of household preferences, the simulation routine first groups households into random sets of three neighbors with each of the 8 geographic regions.  The routine then orders households in the sequential game according to their utilities from using an individual connection.  This assumption captures the intuition that households who would have connected in the absence of any sharing are likely to be the first to take-up a new connection.  Random draws of the preference shocks, $\eta$ and $\epsilon$, are assigned to each household so that households can compute expected utilities over \input{tables/sim_time_periods}months for all possible water source choices.  These shocks are sampled from a multivariate normal distribution taking into account the estimated correlation between sharing partners from Table~\ref{table:correstimates}.  
%\footnote{ Future versions will not condition on observed choices in this way, and instead sample households across sharing relationships. }

@@

\begin{table}
\centering
\caption{Connection Fee Discount: Out-of-Sample Test Results}\label{table:oosestimates}
\input{tables/oos_discount_estimates_group} \\ 
Consumer surplus is measured in PhP/month.
\end{table}

@@

Table~\ref{table:fixedfeediscount} summarizes the impacts of this change on source choices and consumer welfare measured in terms of the difference between current utility and utility from using the outside vendor option as in Section~\ref{section:oostest}.  The decrease in the fixed fee and increase in the tariff both act to decrease the incentives for sharing a connection and leading to a drop of \input{tables/discount_sharing_change_group}percentage points in households using water from their neighbors.  The third row indicates even greater substitution toward water vendors of around \input{tables/discount_vendor_change_group}percentage points in response to the higher tariff.  These effects combine to reduce average consumer surplus by \input{tables/discount_avg_surplus_change_group}PhP per month, which is over half of the initial fixed fee discount of \input{tables/c_fee_discount_group}PhP.  This decline translates into a \input{tables/discount_avg_welfare_change_group}percent loss of monthly household income.  Households in the bottom \input{tables/group_percentile_group}\unskip th percentile of water users also experience lower consumer surplus although the magnitude is much less, amounting to \input{tables/discount_low_welfare_change_group}percent of monthly household income.  Taken together, these results suggest that connection fee discount policies in this context can have unintended distributional consequences by reducing both access to piped water and welfare for the intended beneficiaries. %\footnote{Consumer surplus is measured relative to utility from using vended water.}  
\begin{table}
	\centering
	\caption{Fixed Fee Discount Counterfactual}\label{table:fixedfeediscount}
	\input{tables/discount_estimates_group} \\
	\footnotesize{Low Users include the bottom 50\% of water users. \\
	Surplus is measured in terms of PhPd/month.}
\end{table}

@@

\begin{table}
\centering
\caption{Household Demographics across \\ Survey and Census Data by Water Source}\label{table:pawstocensus}
\input{tables/descriptives_census_to_paws} \\
\footnotesize{The data include connection survey data on households owning water connections \\ as well as 2010 Census data on households.  This sample includes overlapping villages between surveys (67\%)}
\end{table}

@@

\begin{table}
\centering
\caption{Mobile and PAWS Water Source Choices}\label{table:pollfishpawsmeans}
\input{tables/pollfish_paws_means} 
\vspace{.3cm}
\footnotesize{The PAWS data have been adjusted to reflect that \\ shared connections serve multiple households.}
\end{table}

@@

\begin{table}
\caption{Descriptives for Estimation Sample}
\begin{center}
\input{tables/new_descriptives_est_sample}
\end{center}
\end{table}

@@

>>>>

A full analysis of water usage in Manila also requires describing households that choose to use water from local vendors.  For information on households that are using water from alternative sources, I turn to the 2010 Census of Population and Housing.  Households are defined as connected to the piped network if they report using a ``faucet community water system'' for their cooking, laundry, or bathing needs.\footnote{The census was designed largely for rural populations so it does not directly ask whether households are connected to a piped water network.}  Alternatively, I define households as using from a water vendor if they report any other category, which includes deep wells, peddlers, or other sources.  I then construct similar demographic measures to the connection survey using questions in the census and compare these measures across these data sources in Table~\ref{table:pawstocensus} of Appendix~\ref{appendix:pawstocensus}.  Households that own a connection in the connection survey are similar to households connected to the piped water network in the Census data.  To construct a profile of water source choices for the full population, I randomly sample from the pool of households using from vendors in 8 regions of the city so that the data form a representative sample when combined with households covered by the connection survey in each small area.\footnote{Regions are chosen to overlap with ward administrative/political boundaries while ensuring that I am able to obtain a large enough sample of vendor households for each area to perform the counterfactual analysis in Section~\ref{section:fixedcostestimation}.}

@@

With the full set of household preferences, the simulation routine first groups households into random sets of three neighbors with each of the 599 geographic regions.  The routine then orders households in the sequential game according to their utilities from using an individual connection.  This assumption captures the intuition that households who would have connected in the absence of any sharing are likely to be the first to take-up a new connection.  Random draws of the preference shocks, $\eta$ and $\epsilon$, are assigned to each household so that households can compute expected utilities over \input{tables/sim_time_periods}months for all possible water source choices.  These shocks are sampled from a multivariate normal distribution taking into account the estimated correlation between sharing partners from Table~\ref{table:correstimates}.  
%\footnote{ Future versions will not condition on observed choices in this way, and instead sample households across sharing relationships. }

@@

\begin{table}
\centering
\caption{Connection Fee Discount: Out-of-Sample Test Results}\label{table:oosestimates}
\input{tables/oos_discount_estimates_group} \\ 
\footnotesize{Consumer surplus is measured in PhP/month.}
\end{table}

@@

Table~\ref{table:fixedfeediscount} summarizes the impacts of this change on source choices and consumer welfare measured in terms of the difference between current utility and utility from using the outside vendor option as in Section~\ref{section:oostest}.  The decrease in the fixed fee and increase in the tariff both act to decrease the incentives for sharing a connection and leading to a drop of \input{tables/discount_sharing_change_group}percentage points in households using water from their neighbors.  The third row indicates even greater substitution toward water vendors of around \input{tables/discount_vendor_change_group}percentage points in response to the higher tariff.  These effects combine to reduce average consumer surplus by \input{tables/discount_avg_surplus_change_group}PhP per month, which is over half of the initial fixed fee discount of \input{tables/c_fee_discount_group}PhP.  This decline translates into a \input{tables/discount_avg_welfare_change_group}percent loss of monthly household income.  Households in the bottom \input{tables/group_percentile_group}\unskip th percentile of water users also experience lower consumer surplus although the magnitude is much less, amounting to \input{tables/discount_low_welfare_change_group}percent of monthly household income.  Taken together, these results suggest that connection fee discount policies in this context can have unintended distributional consequences by reducing both access to piped water and welfare for the intended beneficiaries. %\footnote{Consumer surplus is measured relative to utility from using vended water.}  
\begin{table}
	\centering
	\caption{Fixed Fee Discount Counterfactual}\label{table:fixedfeediscount}
	\input{tables/discount_estimates_group} \\
	\footnotesize{Low Users include the bottom 50\% of water users. \\
	Surplus is measured in terms of PhP/month.}
\end{table}

@@

\begin{table}
\centering
\caption{Household Demographics across \\ Survey and Census Data by Water Source}\label{table:pawstocensus}
\input{tables/descriptives_census_to_paws} \\
\footnotesize{The data include connection survey data on households owning water connections \\ as well as 2010 Census data on households. This sample includes \\ overlapping wards between  surveys (67\%)}
\end{table}

@@

\begin{table}
\centering
\caption{Mobile and PAWS Water Source Choices}\label{table:pollfishpawsmeans}
\input{tables/pollfish_paws_means} \\
\vspace{.3cm}
\footnotesize{The PAWS data have been adjusted to reflect that \\ shared connections serve multiple households.}
\end{table}

@@

Table~\ref{table:descriptivesestsample} provides descriptive statistics for the connection-level sample used in the simulated method of moments estimation.  Demographic variables are expressed in terms of dummy variables capturing various features that might drive demand for water.  Ward-level attributes include features that determine hassle-costs of sharing with a neighbor.  The sample includes a random sample of 5,000 connections combined with a sample of unconnected households in order to ensure a representative sample of the population of households.

\begin{table}
\caption{Descriptives for Estimation Sample}\label{table:descriptivesestsample}
\begin{center}
\input{tables/new_descriptives_est_sample}
\end{center}
\end{table}

@@

\section{Simulated Method of Moments Model Fit}\label{appendix:modelfit}

By weighting the first three moments highly, the model almost exactly matches moments shown in Table~\ref{table:modelfit}.  The model is also able to match the direction but predicts stronger correlations with fixed preferences for water $\beta_i$ and percent owning connections and percent using from a neighbor.  One interpretation is that there may be other factors besides simply water preferences and heterogeneity in hassle costs that are driving source decisions.  The model has difficulty matching correlation between fixed preferences and share using from vendors, which may be partially due to noise generated in the process of projecting fixed preferences.

\begin{table}
\centering
\caption{Simulated Method of Moments Model Fit}\label{table:modelfit}
\input{tables/smm_moments}
\end{table}
